Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-09-08 src/tex/ms.tex
fix some minor comments

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}

\usepackage{graphicx}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}

@@

\newcommand{\ripple}{\texttt{ripple}}
\newcommand{\python}{\texttt{python}}
\newcommand{\jax}{\texttt{jax}}
\newcommand{\zdethp}{\texttt{zdethp}}
\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\definecolor{cyan}{HTML}{0097A7}
\newcommand{\kl}[1]{{\color{cyan}[KL: #1 ]}}
\definecolor{rr}{RGB}{173, 37, 37}
\newcommand{\te}[1]{{\color{rr}[TE: #1 ]}}

@@

% Begin!
\begin{document}

% Title
\title{Recalibrating Gravitational Wave Phenomenological Waveform Model}

% Author list
\author{Kelvin K.~H.~Lam} 
\email{kelvin33550336@gmail.com}
\cuhk
\author{Kaze W.~K.~Wong} 
\flatiron
\author{Thomas D.~P.~Edwards}
\JHU

@@

% Brief intro of waveform models

Many data analysis tasks in gravitational wave (GW) astrophysics, such
as match filtering~\citep{owen1996search, owen1999matched} and parameter 
estimation~\citep{Dax:2021tsq, Christensen:2022bxb, dynesty, Islam:2022afg, 
Romero-Shaw:2020owr, zackay2018relative, Ashton:2018jfp}, rely upon accurate waveform
models. Since the generation of numerical relativity (NR) waveforms is
prohibitively expensive, the community has constructed waveform approximants
that can be evaluated much faster. There are three families of commonly used GW
approximants: the effective-one-body (EOB) \citep{ossokine2020multipolar,
cotesta2020frequency, taracchini2014effective}, NR
surrogate \citep{islam2022surrogate, varma2019surrogate, varma2019surrogate2},
and phenomenological (Phenom) models \citep{husa2016frequency,
khan2016frequency, garcia2020multimode, pratten2021computationally}. While the
detailed construction of each model is different, they all have a set of
internal parameters that can be calibrated to NR waveforms. The quality of the
waveform model is therefore determined by the ansatz used and the accuracy of the
calibrated parameters.

% Why there is a need

@@

% AD waveforms

@@

% Put more focus on the joint optimization instead of having a set of more accurate 
% coefficients, since the method is the main point
In this paper, we investigate the possibility of further improving the accuracy
of a waveform model, IMRPhenomD \citep{khan2016frequency, husa2016frequency}, by jointly optimizing all the
calibration coefficients for given a set of NR waveforms. Using a similar set of NR
waveforms to those used in \citep{khan2016frequency, husa2016frequency},
we demonstrate that one can improve the match between IMRPhenomD and NR
waveforms over a decently sized parameter space, up to a mass ratio $q=8$. We additionally
explore how different parts of the source parameter space (e.g. the primary and
secondary spins) respond to the optimization procedure by optimizing the waveform
separately for different regions. This can help in understanding
whether the waveform model ansatz performs equally well in different regions of the
parameter space.

@@

% Paragraphs we need:
% 1. What is the model
% 2. How do the parameters enter the model
We start by giving a succinct summary of the IMRPhenomD model and the relevant parameters.
Interested readers should refer to \citep{khan2016frequency} for more details.

Aligned-spin, frequency-domain waveform models (such as IMRPhenomD) can be written as a
combination of amplitude and phase functions ($A$ and $\phi$ respectively):
\begin{align}\label{eq:}
	h(f,\theta,\Lambda) = A(f,\theta,\Lambda)e^{-i\phi(f,\theta,\Lambda)}\,,
\end{align}
where $f$ is the frequency, $\theta$ are the intrinsic parameters of the binary, and $\Lambda$ is a set of additional
parameters which will be discussed below. 
The phase and amplitude functions are then split into three sections which represent the
inspiral, intermediate, and merger-ringdown (MR) parts of the waveform. 
During inspiral, $A$ and $\phi$ are known analytically from post-Newtonian (PN) theory;
IMRPhenomD uses the TaylorF2 model \citep{Buonanno:2009zt, Arun:2004hn} which is known up to 3.5PN order.
To model the intermediate and MR regions, IMRPhenomD (and all waveforms in the IMRPhenom family)
uses a series of parameterizations\footnote{
	The parameterizations for both the amplitude and phase functions can be found in \citep{khan2016frequency}.
} 
which depend purely on $\Lambda$ and can be calibrated to numerical relativity (NR) simulations.
The three sections are then \textit{stitched} together using step functions.
Importantly, the parameterizations are chosen such that they can be made $\mathcal{C}^1$ continuous at the
boundary between each section.

\begin{figure}
	\script{loss.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/loss.pdf}
	\caption{Loss functions against the number of iterations. The vertical
	dotted line indicates the minimum of the validation loss, which is where we stop
	the optimization.}
	\label{fig:loss}
\end{figure}

In practice the $\Lambda$ parameters are fit for each section independently, i.e., intermediate coefficients are fit whilst ignoring the MR region.
Finally, to map the grid of tuned $\Lambda$ parameters back to the intrinsic parameter space, IMRPhenomD uses the polynomial function:
\begin{align} \label{eq:Lambda}
	\Lambda^i&=\lambda_{00}^i+\lambda_{10}^i\eta \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)(\lambda_{01}^i+\lambda_{11}^i\eta+\lambda_{21}^i\eta^2) \nonumber \\ 
	&+(\chi_{\mathrm{PN}}-1)^2(\lambda_{02}^i+\lambda_{12}^i\eta+\lambda_{22}^i\eta^2) \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)^3(\lambda_{03}^i+\lambda_{13}^i\eta+\lambda_{23}^i\eta^2)\,,
\end{align}
where the $\lambda$'s are the fitting coefficients we are going to optimize below, $\eta$ is
the symmetric mass ratio, and $\chi_{\mathrm{PN}}$ is the post-Newtonian spin
parameter, which is defined as 
\begin{align}
	\chi_{\mathrm{PN}}=\frac{m_1\chi_1+m_2\chi_2}{m_1+m_2}-\frac{38\eta}{113}(\chi_1+\chi_2)\,.
\end{align}
Here, $m_{1,2}$ and $\chi_{1,2}$ are the primary and secondary mass and spin,
respectively.

%\kw{Need to work on this paragraph}
Although initially independent, the stitching procedure means that each section
of the waveform intrinsically depends on the full set of $\lambda$'s. 
A slightly inaccurate set of $\lambda$'s can therefore lead to inaccuracies in
the generated waveforms. 
Thus, the calibration of these coefficients is crucial to the accuracy
of IMRPhenom GW models. 
Importantly, since the fitting was performed on the individual segments,
the final waveform is not guaranteed to have $\lambda$'s close to global minima.

@@

In order to optimize the coefficients, we need to define a loss function that
quantifies the difference between the waveform model and the target NR simulations which we want to match.
Here we adopt a quantity commonly used in GW physics called the \textit{mismatch} function \citep{owen1996search, husa2016frequency}. 
It is defined as
\begin{align} \label{eq:mismatch}
	\mathcal{M}(h_1, h_2)=1-\max_{t_0, \phi_0}\langle \hat{h}_1, \hat{h}_2\rangle,
\end{align}
where $h_{1,2}$ are the two GW waveforms we are comparing, and $t_0$ and $\phi_0$
are a relative time and phase shift respectively. 
The inner product, $\langle h_1, h_2 \rangle$, is defined as 
\begin{align}\label{eq:inner_prod}
	\langle h_1, h_2 \rangle = 4\Re\int_{f_{\mathrm{min}}}^{f_{\mathrm{max}}}\frac{h_1(f)h_2^{\ast}(f)}{S_n(f)}\,df,
\end{align}
where $\hat{h}=h/\sqrt{\langle h, h \rangle}$ is the normalized GW strain,
$S_n(f)$ is the power spectral density (PSD), and $f_{\mathrm{max}}$ ($f_{\mathrm{min}}$) is
the maximum (minimum) frequency for the integration.
We note here that the mismatch can be viewed as the mean square error (MSE) between the
two waveforms.

@@

Since we wish to optimize the model over the whole parameter space,
we need to compare multiple model generated waveforms with NR waveforms.
However, the mismatch is only defined for two input waveforms at a particular set of intrinsic parameters.
We therefore define the loss function as an average of training waveforms in two ways,
the simple average of mismatches and the normalized average of mismatches,  
\begin{align}\label{eqn:loss}
	\mathcal{L}_{\mathrm{ave}}&=\frac{1}{N}\sum_{i=1}^N\mathcal{M}_i \\
	\mathcal{L}_{\mathrm{norm}}&=\frac{1}{N}\sum_{i=1}^N\frac{\mathcal{M}_i}{\mathcal{M}_{i,\mathrm{ini}}},
\end{align}	
where $\mathcal{M}_i$ represents the mismatch of an individual training waveform,
$\mathcal{M}_{i,\mathrm{ini}}$ represents the initial mismatch of the individual
training waveform, and $N$ is the total number of training waveforms.

The two loss functions are chosen to give different 
behavior during the optimization process.
The simple average, $\mathcal{L}_{\mathrm{mean}}$, serves as the simplest choice of loss function,
but is prone to be dominated by a single point in parameter space with a large mismatch.
Other points with smaller mismatches would be insignificant comparatively, and might not be able 
to improve under such a loss function.
Alternatively, the normalized average, $\mathcal{L}_{\mathrm{norm}}$, eliminates the aforementioned issue
by encouraging the waveform to improve at each training point at a similar rate.
The ratio in $\mathcal{L}_{\mathrm{norm}}$ will therefore remain approximately the same for each training point.
Conversely, $\mathcal{L}_{\mathrm{mean}}$ allows the loss function to automatically adjust
and preferentially optimize the largest mismatches, encouraging the waveform to have similar mismatches everywhere.
In this paper, we show the results of using both loss functions 
and examine the differences between them.

@@

The SXS NR waveforms are given as time-series strain.
Since IMRPhenomD is modeled in the frequency domain, we need to Fourier transform the simulation results
in order to compute the mismatch, \eqref{eq:mismatch}.
For this, we taper the time-series using Tukey window~\citep{usman2016pycbc}\footnote{
	Specifically, we choose $\alpha=2t_{\mathrm{RD}}/T$, where
	$t_{\mathrm{RD}}$ is the duration of ringdown (maximum amplitude to the end of the strain) and $T$ is the duration of the
	entire GW strain. 
%	\te{What is $t_{\mathrm{RD}}$? and how is it computed?}
} before using standard FFT routines to compute the fast Fourier transform.

@@

\begin{algorithm}[t]
	\caption{Gradient descent pseudocode}
	\label{alg:gradient}
	\KwIn{initial coefficients $\lambda_i$} \Parameters{number of iterations
		$N$, learning rate $\alpha$} \Variables{current coefficients $\lambda$,
		mismatch gradient $\nabla\mathcal{L}$} \KwResult{output coefficients
		$\lambda$} $\lambda\leftarrow\lambda_i$\\
	\tcc{Gradient Descent}
	\For{$i<N$}{ $\mathcal{L}\leftarrow Mismatch(\lambda)$ \\
		$\nabla\mathcal{L}\leftarrow AutoDiff(\mathcal{L})$\\
		$\lambda\leftarrow\lambda-\alpha\nabla\mathcal{L}$\\
	} \Return{$\lambda$}
\end{algorithm}

@@

\begin{figure}[t]
	\script{q148.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148.pdf}
	\caption{Distributions of mismatches before and after optimization using the 
	$\mathcal{L}_{\mathrm{ave}}$ loss function. Mismatches are calculated using the training waveforms listed in 
	Tab.~\ref{tab:q148} and	are weighted with a constant PSD. 
	The dashed lines represent the median of the distributions, which decreased by 45.3\% during optimization.}
	\label{fig:q148}
\end{figure}
\begin{figure}[t]
	\script{q148_q1248_compare.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148_q1248_compare.pdf}
	\caption{Distributions of $\log_{10}$ difference in mismatch. The red
	distribution uses the training waveforms listed in
	Tab.~\ref{tab:q148} while the green distribution uses waveforms
	listed in Tab.~\ref{tab:q148}~and~\ref{tab:q1248}. Mismatches are calculated
	using a constant PSD with the loss function $\mathcal{L}_{\mathrm{ave}}$. 
	The dashed lines represent the median of the distributions, which decreased by an additional 10.8\% during optimization with the additional data.}
	\label{fig:q148_q1248_compare}
\end{figure}

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately
(Fig.~\ref{fig:all_quadrants}~and~\ref{fig:ps_q148_quadrant}). The training waveforms used for fitting in
this scenario are listed in Tab.~\ref{tab:quadrants} and the loss functions 
are calculated using a simple average of the mismatches, $\mathcal{L}_{\mathrm{ave}}$. The top-left and
bottom-right regions have limited data for $q>4$, hence the result is only valid up to $q\leq4$ and we only use waveforms with $q\leq4$ as validation waveforms for these two regions. 
From Fig.~\ref{fig:all_quadrants}, we observe that both the top-right and bottom-left regions improve significantly over the original model.
This is especially pronounced for the bottom-left region, where the improvement is significantly better than optimizing all regions simultaneously.
This suggests the ansatz fits this part of the parameter space well.
The top-left region also improves over the original model although it is similar to when optimizing all waveforms at once.
% Notice optimizing all regions gives a slightly better improvement.
% This shows that the ansatz can achieve a better fit using training waveforms with similar spins, even for waveforms with non-equal spin parameters.}
On the other hand, the bottom-right region does not improve over the original model, indicating that a change to the ansatz is required to fit the NR data better.
% Interestingly, we can see in the bottom-right region that the waveform does not show a significant change
% to the overall mismatch (see bottom-right panel of Fig.~\ref{fig:all_quadrants}). 
% This suggests that the IMRPhenomD waveform ansatz fits waveforms with
% $\chi_1>0$ and $\chi_2<0$ less well.

@@

\begin{figure*}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\textwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches for both split region optimization and all region optimization in all 4 
	regions. We use a constant PSD to calculate the mismatch and
	$\mathcal{L}_{\mathrm{ave}}$ as the loss function. The medians of each distribution are indicated by dashed
	lines.}
	\label{fig:all_quadrants}
\end{figure*}

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{Combined mismatch distributions for both optimization schemes shown in Fig.~\ref{fig:all_quadrants}.}
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

While this study focuses on the IMRPhenomD model due to its availability as an
automatically differentiable waveform in \ripple, other more sophisticated waveforms such as
IMRPhenomP \citep{hannam2014simple, khan2019phenomenological} and IMRPhenomXAS
\citep{pratten2020setting,pratten2021computationally} can also potentially benefit
from recalibration. In fact, since these waveforms have a larger number of calibration
parameters and are fit to a larger number of NR simulations, it is possible that the 
improvement will be more dramatic than for IMRPhenomD. {\ripple} is being regularly
updated with new waveforms, and future work should carefully examine whether these more
modern waveforms can be improved. % Extend this a bit?

@@

We thank Will M.~Farr, Max Isi, and Mark Hannam for helpful discussions; we also
thank Carl-Johan Haster, Neil J.~Cornish and Thomas Dent for comments on the
draft. The Flatiron Institute is a division of the Simons Foundation. T.E.\ is
supported by the Horizon Postdoctoral Fellowship.
\bibliography{bib}

>>>>

\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{graphicx}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}

@@

\newcommand{\ripple}{\texttt{ripple}}
\newcommand{\python}{\texttt{python}}
\newcommand{\jax}{\texttt{jax}}
\newcommand{\zdethp}{\texttt{zdethp}}

\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\definecolor{cyan}{HTML}{0097A7}
\newcommand{\kl}[1]{{\color{cyan}[KL: #1 ]}}
\definecolor{rr}{RGB}{173, 37, 37}
\newcommand{\te}[1]{{\color{rr}[TE: #1 ]}}
\definecolor{cyan}{HTML}{0097A7}
\newcommand{\rr}[1]{{\color{cyan}[#1]}}

@@

\begin{document}

\title{Recalibrating Gravitational Wave Phenomenological Waveform Model}

\author{Kelvin K.~H.~Lam} 
\email{kelvin33550336@gmail.com}
\cuhk
\author{Kaze W.~K.~Wong} 
\flatiron
\author{Thomas D.~P.~Edwards}
\JHU

@@

Many data analysis tasks in gravitational wave (GW) astrophysics, such
as \rr{matched} filtering~\citep{owen1996search, owen1999matched} and parameter 
estimation~\citep{Dax:2021tsq, Christensen:2022bxb, dynesty, Islam:2022afg, 
Romero-Shaw:2020owr, zackay2018relative, Ashton:2018jfp}, rely upon accurate waveform
models. Since the generation of numerical relativity (NR) waveforms is
prohibitively expensive, the community has constructed waveform approximants
that can be evaluated much faster. There are three families of commonly used GW
approximants: the effective-one-body (EOB) \citep{ossokine2020multipolar,
cotesta2020frequency, taracchini2014effective}, NR
surrogate \citep{islam2022surrogate, varma2019surrogate, varma2019surrogate2},
and phenomenological (Phenom) models \citep{husa2016frequency,
khan2016frequency, garcia2020multimode, pratten2021computationally}. While the
detailed construction of each model is different, they all have a set of
internal parameters that can be calibrated to NR waveforms. The quality of the
waveform model is therefore determined by the ansatz used and the accuracy of the
calibrated parameters.

@@

@@

In this paper, we investigate the possibility of further improving the accuracy
of a waveform model, IMRPhenomD \citep{khan2016frequency, husa2016frequency}, by jointly optimizing all the
calibration coefficients for given a set of NR waveforms. Using a similar set of NR
waveforms to those used in \citep{khan2016frequency, husa2016frequency},
we demonstrate that one can improve the match between IMRPhenomD and NR
waveforms over a decently sized parameter space, up to a mass ratio $q=8$. We additionally
explore how different parts of the source parameter space (e.g. the primary and
secondary spins) respond to the optimization procedure by optimizing the waveform
separately for different regions. This can help in understanding
whether the waveform model ansatz performs equally well in different regions of the
parameter space.

@@

We start by giving a succinct summary of the IMRPhenomD model and the relevant parameters.
Interested readers should refer to \citep{khan2016frequency} for more details.

Aligned-spin, frequency-domain waveform models (such as IMRPhenomD) can be written as a
combination of amplitude and phase functions ($A$ and $\phi$ respectively):
\begin{align}\label{eq:}
	h(f,\theta,\Lambda) = A(f,\theta,\Lambda)e^{-i\phi(f,\theta,\Lambda^i)}\,,
\end{align}
where $f$ is the frequency, $\theta$ are the intrinsic parameters of the binary, and $\Lambda^i$ \rr{($i=1,2,\cdots,19$) is a set of 19 additional
parameters} which will be discussed below. 
The phase and amplitude functions are then split into three sections which represent the
inspiral, intermediate, and merger-ringdown (MR) parts of the waveform. 
During inspiral, $A$ and $\phi$ are known analytically from post-Newtonian (PN) theory;
IMRPhenomD uses the TaylorF2 model \citep{Buonanno:2009zt, Arun:2004hn} which is known up to 3.5PN order.
To model the intermediate and MR regions, IMRPhenomD (and all waveforms in the IMRPhenom family)
uses a series of parameterizations\footnote{
	The parameterizations for both the amplitude and phase functions can be found in \citep{khan2016frequency}.
} 
which depend purely on $\Lambda^i$ and can be calibrated to numerical relativity (NR) simulations.
The three sections are then \textit{stitched} together using step functions.
Importantly, the parameterizations are chosen such that they can be made $\mathcal{C}^1$ continuous at the
boundary between each section.

\begin{figure}
	\script{loss.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/loss.pdf}
	\caption{\rr{Average Loss $\mathcal{L}_{\mathrm{ave}}$} against the number of iterations. The vertical
	dotted line indicates the minimum of the validation loss, which is where we stop
	the optimization.}
	\label{fig:loss}
\end{figure}

In practice the $\Lambda^i$ parameters are fit for each section independently, i.e., intermediate coefficients are fit whilst ignoring the MR region.
Finally, to map the grid of tuned $\Lambda^i$ parameters back to the intrinsic parameter space, IMRPhenomD uses the polynomial function:
\begin{align} \label{eq:Lambda}
	\Lambda^i&=\lambda_{00}^i+\lambda_{10}^i\eta \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)(\lambda_{01}^i+\lambda_{11}^i\eta+\lambda_{21}^i\eta^2) \nonumber \\ 
	&+(\chi_{\mathrm{PN}}-1)^2(\lambda_{02}^i+\lambda_{12}^i\eta+\lambda_{22}^i\eta^2) \nonumber \\
	&+(\chi_{\mathrm{PN}}-1)^3(\lambda_{03}^i+\lambda_{13}^i\eta+\lambda_{23}^i\eta^2)\,,
\end{align}
where the $\lambda$'s are the fitting coefficients we are going to optimize below, $\eta$ is
the symmetric mass ratio, and $\chi_{\mathrm{PN}}$ is the post-Newtonian spin
parameter, which is defined as 
\begin{align}
	\chi_{\mathrm{PN}}=\frac{m_1\chi_1+m_2\chi_2}{m_1+m_2}-\frac{38\eta}{113}(\chi_1+\chi_2)\,.
\end{align}
Here, $m_{1,2}$ and $\chi_{1,2}$ are the primary and secondary mass and spin,
respectively.

Although initially independent, the stitching procedure means that each section
of the waveform intrinsically depends on the full set of $\lambda$'s. 
A slightly inaccurate set of $\lambda$'s can therefore lead to inaccuracies in
the generated waveforms. 
Thus, the calibration of these coefficients is crucial to the accuracy
of IMRPhenom GW models. 
Importantly, since the fitting was performed on the individual segments,
the final waveform is not guaranteed to have $\lambda$'s close to global minima.

@@

In order to optimize the coefficients, we need to define a loss function that
quantifies the difference between the waveform model and the target NR simulations which we want to match.
Here we adopt a quantity commonly used in GW physics called the \textit{mismatch} function \citep{owen1996search, husa2016frequency}. 
It is defined as
\begin{align} \label{eq:mismatch}
	\mathcal{M}(h_1, h_2)=1-\max_{t_0, \phi_0}\langle \hat{h}_1, \hat{h}_2\rangle,
\end{align}
where $h_{1,2}$ are the two GW waveforms we are comparing, and $t_0$ and $\phi_0$
are a relative time and phase shift respectively. 
The inner product, $\langle h_1, h_2 \rangle$, is defined as 
\begin{align}\label{eq:inner_prod}
	\langle h_1, h_2 \rangle = 4\Re\int_{f_{\mathrm{min}}}^{f_{\mathrm{max}}}\frac{h_1(f)h_2^{\ast}(f)}{S_n(f)}\,df,
\end{align}
where $\hat{h}=h/\sqrt{\langle h, h \rangle}$ is the normalized GW strain,
$S_n(f)$ is the power spectral density (PSD), and $f_{\mathrm{max}}$ ($f_{\mathrm{min}}$) is
the maximum (minimum) frequency for the integration.
We note here that the mismatch can be viewed as \rr{a} mean square error (MSE) between the
two waveforms.

@@

Since we wish to optimize the model over the whole parameter space,
we need to compare multiple model generated waveforms with NR waveforms.
However, the mismatch is only defined for two input waveforms at a particular set of intrinsic parameters.
We therefore define the loss function as an average of training waveforms in two ways,
the simple average of mismatches and the normalized average of mismatches,  
\begin{align}\label{eqn:loss}
	\mathcal{L}_{\mathrm{ave}}&=\frac{1}{N}\sum_{i=1}^N\mathcal{M}_i \\
	\mathcal{L}_{\mathrm{norm}}&=\frac{1}{N}\sum_{i=1}^N\frac{\mathcal{M}_i}{\mathcal{M}_{i,\mathrm{ini}}},
\end{align}	
where $\mathcal{M}_i$ represents the mismatch of an individual training waveform,
$\mathcal{M}_{i,\mathrm{ini}}$ represents the mismatch \rr{between original IMRPhenomD waveforms and NR waveforms}, and $N$ is the total number of training waveforms.

The two loss functions are chosen to give different 
behavior during the optimization process.
The simple average, \rr{$\mathcal{L}_{\mathrm{ave}}$}, serves as the simplest choice of loss function,
but is prone to be dominated by a single point in parameter space with a large mismatch.
Other points with smaller mismatches would be insignificant comparatively, and might not be able 
to improve under such a loss function.
Alternatively, the normalized average, $\mathcal{L}_{\mathrm{norm}}$, eliminates the aforementioned issue
by encouraging the waveform to improve at each training point at a similar rate.
The ratio in $\mathcal{L}_{\mathrm{norm}}$ will therefore remain approximately the same for each training point.
Conversely, \rr{$\mathcal{L}_{\mathrm{ave}}$} allows the loss function to automatically adjust
and preferentially optimize the largest mismatches, encouraging the waveform to have similar mismatches everywhere.
In this paper, we show the results of using both loss functions 
and examine the differences between them.

@@

The SXS NR waveforms are given as time-series strain.
Since IMRPhenomD is modeled in the frequency domain, we need to Fourier transform the simulation results
in order to compute the mismatch, \eqref{eq:mismatch}.
For this, we taper the time-series using Tukey window \citep{usman2016pycbc}
\footnote{
	Specifically, we choose $\alpha=2t_{\mathrm{RD}}/T$, where
	$t_{\mathrm{RD}}$ is the duration of ringdown (maximum amplitude to the end of the strain) and $T$ is the duration of the
	entire GW strain. 
} 
before using standard FFT routines to compute the fast Fourier transform.

@@

\begin{algorithm}[t]
	\caption{Gradient descent pseudocode}
	\label{alg:gradient}
	\KwIn{initial coefficients $\lambda_i$} \Parameters{number of iterations
		$M$, learning rate $\alpha$} \Variables{current coefficients $\lambda$,
		mismatch gradient $\nabla\mathcal{L}$} \KwResult{output coefficients
		$\lambda$} $\lambda\leftarrow\lambda_i$\\
	\tcc{Gradient Descent}
	\For{$i<M$}{ $\mathcal{L}\leftarrow Mismatch(\lambda)$ \\
		$\nabla\mathcal{L}\leftarrow AutoDiff(\mathcal{L})$\\
		$\lambda\leftarrow\lambda-\alpha\nabla\mathcal{L}$\\
	} \Return{$\lambda$}
\end{algorithm}

@@

\begin{figure}[t]
	\script{q148.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148.pdf}
	\caption{Distributions of mismatches before and after optimization using the 
	$\mathcal{L}_{\mathrm{ave}}$ loss function. Mismatches are calculated using the training waveforms listed in 
	Tab.~\ref{tab:q148} and	are weighted with a constant PSD. 
	The dotted lines represent the median of the distributions, which decreased by 45.3\% during optimization.}
	\label{fig:q148}
\end{figure}
\begin{figure}[t]
	\script{q148_q1248_compare.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148_q1248_compare.pdf}
	\caption{Distributions of $\log_{10}$ difference in mismatch. The red
	distribution uses the training waveforms listed in
	Tab.~\ref{tab:q148} while the green distribution uses waveforms
	listed in Tab.~\ref{tab:q148}~and~\ref{tab:q1248}. Mismatches are calculated
	using a constant PSD with the loss function $\mathcal{L}_{\mathrm{ave}}$. 
	The dotted lines represent the median of the distributions, which decreased by an additional 10.8\% during optimization with the additional data.}
	\label{fig:q148_q1248_compare}
\end{figure}

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately
(Fig.~\ref{fig:all_quadrants}~and~\ref{fig:ps_q148_quadrant}). The training waveforms used for fitting in
this scenario are listed in Tab.~\ref{tab:quadrants} and the loss functions 
are calculated using a simple average of the mismatches, $\mathcal{L}_{\mathrm{ave}}$. The top-left and
bottom-right regions have limited data for $q>4$, hence the result is only valid up to $q\leq4$ and we only use waveforms with $q\leq4$ as validation waveforms for these two regions. 
From Fig.~\ref{fig:all_quadrants}, we observe that both the top-right and bottom-left regions improve significantly over the original model.
This is especially pronounced for the bottom-left region, where the improvement is significantly better than optimizing all regions simultaneously.
This suggests the ansatz fits this part of the parameter space well.
The top-left region also improves over the original model although it is similar to when optimizing all waveforms at once.
On the other hand, the bottom-right region does not improve over the original model, indicating that a change to the ansatz is required to fit the NR data better.

@@

\begin{figure*}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\textwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches for both split region optimization and all region optimization in all 4 
	regions. We use a constant PSD to calculate the mismatch and
	$\mathcal{L}_{\mathrm{ave}}$ as the loss function. The medians of each distribution are indicated by dotted
	lines.}
	\label{fig:all_quadrants}
\end{figure*}

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{Combined mismatch distributions for both optimization schemes shown in Fig.~\ref{fig:all_quadrants}. The medians of both distributions are indicated by dotted lines. }
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

While this study focuses on the IMRPhenomD model due to its availability as an
automatically differentiable waveform in \ripple, other more sophisticated waveforms such as
IMRPhenomP \citep{hannam2014simple, khan2019phenomenological} and IMRPhenomXAS
\citep{pratten2020setting,pratten2021computationally} can also potentially benefit
from recalibration. In fact, since these waveforms have a larger number of calibration
parameters and are fit to a larger number of NR simulations, it is possible that the 
improvement will be more dramatic than for IMRPhenomD. {\ripple} is being regularly
updated with new waveforms, and future work should carefully examine whether these more
modern waveforms can be improved.

@@

We thank Will M.~Farr, Max Isi, and Mark Hannam for helpful discussions; we also
thank Carl-Johan Haster, Neil J.~Cornish and Thomas Dent for comments on the
draft. The Flatiron Institute is a division of the Simons Foundation. T.E.\ is
supported by the Horizon Postdoctoral Fellowship.

\bibliography{bib}
